---
title: "Factors determining average life expectancy."
author: "Marcel Florjański, Filip Kasprzycki, Bartłomiej Klajn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```
In this project, we will analyze various factors in terms of their impact on life expectancy.
Our dependent variable is:
-life expectancy, 
and our explanatory variables are:
-access to clean fuels and technologies for cooking,
-children per woman,
-gdp per capita,
-share of adults defined as obese,
-share of adults who smoke,
-share of individuals using the internet,
-share of the population with access to electocity.

In this project we will consider 3 main hypotheses:

H1: Economic development (measured by GDP per capita) has a positive and diminishing marginal impact on life expectancy.

H2: Lifestyle risks, specifically smoking and obesity rates, are expected to have a significant negative impact on longevity.

H3: Infrastructure indicators, such as access to electricity and clean fuels, are vital determinants, especially in developing nations.

All the data we use comes from Our World in Data, which is a reputable source for global statistics and research. The data is collected from various sources, including international organizations, research institutions, and government agencies. It is regularly updated and covers a wide range of topics, making it a valuable resource for our analysis of life expectancy and its determinants.

Libraries we are going to use:
```{r}
library(tidyverse)
library(mice)
library(countrycode)
library(corrplot)
library(ggplot2)
```

## 2. Data Loading and Merging
In this section, we load the raw datasets and merge them based on country entity and code for the year 2022.

```{r data_merge_final}
# 1. Loading the Files
y_data   <- read_csv("life-expectancy.csv")
x_gdp    <- read_csv("gdp-per-capita-worldbank.csv")
x_fert   <- read_csv("children-per-woman-un.csv")
x_obese  <- read_csv("share-of-adults-defined-as-obese.csv")
x_smoke  <- read_csv("share-of-adults-who-smoke.csv")
x_net    <- read_csv("share-of-individuals-using-the-internet.csv")
x_elec   <- read_csv("share-of-the-population-with-access-to-electricity.csv")
x_fuel   <- read_csv("access-to-clean-fuels-and-technologies-for-cooking.csv")

# 2. Merging the datasets
df_final <- y_data %>%
  select(Entity, Code, LifeExp = 4) %>%
  inner_join(x_gdp %>% select(Code, GDP = 4), by = "Code") %>%
  inner_join(x_fert %>% select(Code, Fertility = 4), by = "Code") %>%
  inner_join(x_obese %>% select(Code, Obesity = 4), by = "Code") %>%
  inner_join(x_smoke %>% select(Code, Smoking = 4), by = "Code") %>%
  inner_join(x_net %>% select(Code, Internet = 4), by = "Code") %>%
  inner_join(x_elec %>% select(Code, Electricity = 4), by = "Code") %>%
  inner_join(x_fuel %>% select(Code, CleanFuels = 4), by = "Code") %>%
  drop_na()
# 3. Adding continent information using countrycode package
df_final$Continent <- countrycode(sourcevar = df_final$Code,
                                  origin = "iso3c",
                                  destination = "continent")

# 4. Converting Continent to a factor variable
df_final$Continent <- as.factor(df_final$Continent)

table(df_final$Continent)

# 5. Checking the final dataset
nrow(df_final)
head(df_final)
md.pattern(df_final, plot = FALSE)
```

# 3. Descriptive Statistics and Visualization
In this section, we will explore the descriptive statistics of our dataset.

At first, we will look at the summary statistics of our variables to understand their central tendencies and dispersions.
```{r}
summary(df_final %>% select(-Entity, -Code))
```


Next, we will visualize the structure of our variables.
```{r}
hist(df_final$LifeExp, main="Distribution of Life Expectancy", col="lightblue", border="white", xlab="Years")
```


The distribution of life expectancy is slightly skewed to the left, but is strongly concentrated around the ages of 70-80. No outliers.
```{r}
hist(df_final$GDP, main="Distribution of Raw GDP per Capita", col="orange", border="white", xlab="USD")
```


The distribution is very skewed to the right. Most countries have rather low scores (below 20,000 USD), but there are a few that reach over 100,000 USD. For correct modeling, it is advisable to take the logarithm.
```{r}
par(mfrow=c(1,2))
hist(df_final$Obesity, main="Obesity Rates", col="tomato", border="white")
hist(df_final$Smoking, main="Smoking Rates", col="grey", border="white")
par(mfrow=c(1,1))
```


Obesity has a distribution close to normal. Tobacco has a fairly even distribution with a slight right-skew.
```{r}
par(mfrow=c(1,3))
hist(df_final$Internet, main="Internet Access", col="lightgreen", border="white")
hist(df_final$Electricity, main="Electricity Access", col="gold", border="white")
hist(df_final$CleanFuels, main="Clean Fuels Access", col="skyblue", border="white")
par(mfrow=c(1,1))
```


Infrastructure variables are highly specific, with a large number of countries scoring close to 100%. This may suggest collinearity between these variables and GDP levels.



Based on the previous histogram analysis, we apply a natural logarithm to GDP per capita. This transformation is intended to linearize the relationship between income and life expectancy, as suggested by Hypothesis 1.

```{r log_gdp_transform}
df_final$log_GDP <- log(df_final$GDP)


par(mfrow=c(1,2))
plot(df_final$GDP, df_final$LifeExp, main="Raw GDP vs LifeExp", 
     xlab="GDP", ylab="Life Expectancy", pch=19, col=rgb(0.2,0.5,0.5,0.5))
plot(df_final$log_GDP, df_final$LifeExp, main="Log(GDP) vs LifeExp", 
     xlab="log(GDP)", ylab="Life Expectancy", pch=19, col=rgb(0.8,0.2,0.2,0.5))
par(mfrow=c(1,1))
```


After analyzing the graphs, the assumption is confirmed: the relationship between GDP and life expectancy is nonlinear. The effect of diminishing returns is visible: in low-income countries, each additional dollar significantly increases life expectancy, but in wealthy countries, this effect disappears. Using the logarithm "straightened" the graph. For further analysis, we will use the logarithm variable.


Next, we will look at the correlation matrix to understand the relationships between our variables.
```{r}
numeric_data <- df_final %>% select(where(is.numeric))

M <- cor(numeric_data)

corrplot(M, 
         method = "color", 
         addCoef.col = "black",
         number.cex = 0.7, 
         tl.col = "black", 
         tl.srt = 45, 
         diag = FALSE)
```


The correlation matrix confirms the improved effectiveness of the GDP variable by taking it into logarithms – an increase from 0.74 to 0.87. An interesting relationship is the number of children a woman bears and life expectancy – having many children indicates a shorter lifespan, but this is most likely conditioned by the socio-economic situation – having many children is most common in developing countries, where people statistically have shorter lifespans. However, the matrix revealed a strong correlation between the infrastructure variables (both among themselves and in relation to GDP). This may indicate collinearity. This compels us to conduct detailed analysis in this area in the future, particularly the use of VIF in the modeling stage.


At the end for this section we are going to analyze how life expectancy varies across continents.
```{r}
boxplot(LifeExp ~ Continent, data = df_final, 
        col = c("gold", "tomato", "skyblue", "lightgreen", "violet"),
        main = "Life Expectancy by Continent",
        ylab = "Years", xlab = "Continent")


ggplot(df_final, aes(x = log_GDP, y = LifeExp, color = Continent)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(title = "Life Expectancy vs Log(GDP) by Continent")
```


Clear differences between continents are visible. Europe clearly dominates with the highest and highly concentrated results. Africa fares by far the worst. Africa and Asia have very large whiskers, indicating their strong internal differentiation. Analysis of the graphs indicates that the "Continent" variable must be included in the model as a quantitative variable. Without it, the model would distort the results because it would not know that belonging to Europe statistically increases longevity.

#Model preparation- Train/Test split


After the initial analysis of variables, we divide the data into two sets - we will train our model on one set, and then test its effectiveness on the "new" one. We divide the data in an 80/20 ratio.

```{r}
set.seed(123)

train_indices <- sample(1:nrow(df_final), size = 0.8 * nrow(df_final))

train_data <- df_final[train_indices, ]
test_data  <- df_final[-train_indices, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size: ", nrow(test_data))
```