---
title: "Factors determining average life expectancy."
author: "Marcel Florjański, Filip Kasprzycki, Bartłomiej Klajn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```
In this project, we will analyze various factors in terms of their impact on life expectancy.
Our dependent variable is:
-life expectancy, 
and our explanatory variables are:
-access to clean fuels and technologies for cooking,
-children per woman,
-gdp per capita,
-share of adults defined as obese,
-share of adults who smoke,
-share of individuals using the internet,
-share of the population with access to electocity.

In this project we will consider 3 main hypotheses:

H1: Economic development (measured by GDP per capita) has a positive and diminishing marginal impact on life expectancy.

H2: Lifestyle risks, specifically smoking and obesity rates, are expected to have a significant negative impact on longevity.

H3: Infrastructure indicators, such as access to electricity and clean fuels, are vital determinants, especially in developing nations.

All the data we use comes from Our World in Data, which is a reputable source for global statistics and research. The data is collected from various sources, including international organizations, research institutions, and government agencies. It is regularly updated and covers a wide range of topics, making it a valuable resource for our analysis of life expectancy and its determinants.

Libraries we are going to use:
```{r}
library(tidyverse)
library(mice)
library(countrycode)
library(corrplot)
library(ggplot2)
library(car)
library(tseries)
library(lmtest)
```

## 2. Data Loading and Merging
In this section, we load the raw datasets and merge them based on country entity and code for the year 2022.

```{r data_merge_final}
# 1. Loading the Files
y_data   <- read_csv("life-expectancy.csv")
x_gdp    <- read_csv("gdp-per-capita-worldbank.csv")
x_fert   <- read_csv("children-per-woman-un.csv")
x_obese  <- read_csv("share-of-adults-defined-as-obese.csv")
x_smoke  <- read_csv("share-of-adults-who-smoke.csv")
x_net    <- read_csv("share-of-individuals-using-the-internet.csv")
x_elec   <- read_csv("share-of-the-population-with-access-to-electricity.csv")
x_fuel   <- read_csv("access-to-clean-fuels-and-technologies-for-cooking.csv")

# 2. Merging the datasets
df_final <- y_data %>%
  select(Entity, Code, LifeExp = 4) %>%
  inner_join(x_gdp %>% select(Code, GDP = 4), by = "Code") %>%
  inner_join(x_fert %>% select(Code, Fertility = 4), by = "Code") %>%
  inner_join(x_obese %>% select(Code, Obesity = 4), by = "Code") %>%
  inner_join(x_smoke %>% select(Code, Smoking = 4), by = "Code") %>%
  inner_join(x_net %>% select(Code, Internet = 4), by = "Code") %>%
  inner_join(x_elec %>% select(Code, Electricity = 4), by = "Code") %>%
  inner_join(x_fuel %>% select(Code, CleanFuels = 4), by = "Code") %>%
  drop_na()
# 3. Adding continent information using countrycode package
df_final$Continent <- countrycode(sourcevar = df_final$Code,
                                  origin = "iso3c",
                                  destination = "continent")

# 4. Converting char data to a factor variable
df_final$Continent <- as.factor(df_final$Continent)
df_final$Entity <- as.factor(df_final$Entity)
df_final$Code <- as.factor(df_final$Code)
table(df_final$Continent)

# 5. Checking the final dataset
nrow(df_final)
head(df_final)
md.pattern(df_final, plot = FALSE)
```

# 3. Descriptive Statistics and Visualization
In this section, we will explore the descriptive statistics of our dataset.

At first, we will look at the summary statistics of our variables to understand their central tendencies and dispersions.
```{r}
summary(df_final %>% select(-Entity, -Code))
```


Next, we will visualize the structure of our variables.
```{r}
hist(df_final$LifeExp, main="Distribution of Life Expectancy", col="lightblue", border="white", xlab="Years")
```


The distribution of life expectancy is slightly skewed to the left, but is strongly concentrated around the ages of 70-80. No outliers.
```{r}
hist(df_final$GDP, main="Distribution of Raw GDP per Capita", col="orange", border="white", xlab="USD")
```


The distribution is very skewed to the right. Most countries have rather low scores (below 20,000 USD), but there are a few that reach over 100,000 USD. For correct modeling, it is advisable to take the logarithm.
```{r}
par(mfrow=c(1,2))
hist(df_final$Obesity, main="Obesity Rates", col="tomato", border="white")
hist(df_final$Smoking, main="Smoking Rates", col="grey", border="white")
par(mfrow=c(1,1))
```


Obesity has a distribution close to normal. Tobacco has a fairly even distribution with a slight right-skew.
```{r}
par(mfrow=c(1,3))
hist(df_final$Internet, main="Internet Access", col="lightgreen", border="white")
hist(df_final$Electricity, main="Electricity Access", col="gold", border="white")
hist(df_final$CleanFuels, main="Clean Fuels Access", col="skyblue", border="white")
par(mfrow=c(1,1))
```


Infrastructure variables are highly specific, with a large number of countries scoring close to 100%. This may suggest collinearity between these variables and GDP levels.



Based on the previous histogram analysis, we apply a natural logarithm to GDP per capita. This transformation is intended to linearize the relationship between income and life expectancy, as suggested by Hypothesis 1.

```{r log_gdp_transform}
df_final$log_GDP <- log(df_final$GDP)


par(mfrow=c(1,2))
plot(df_final$GDP, df_final$LifeExp, main="Raw GDP vs LifeExp", 
     xlab="GDP", ylab="Life Expectancy", pch=19, col=rgb(0.2,0.5,0.5,0.5))
plot(df_final$log_GDP, df_final$LifeExp, main="Log(GDP) vs LifeExp", 
     xlab="log(GDP)", ylab="Life Expectancy", pch=19, col=rgb(0.8,0.2,0.2,0.5))
par(mfrow=c(1,1))
```


After analyzing the graphs, the assumption is confirmed: the relationship between GDP and life expectancy is nonlinear. The effect of diminishing returns is visible: in low-income countries, each additional dollar significantly increases life expectancy, but in wealthy countries, this effect disappears. Using the logarithm "straightened" the graph. For further analysis, we will use the logarithm variable.


Next, we will look at the correlation matrix to understand the relationships between our variables.
```{r}
numeric_data <- df_final %>% select(where(is.numeric))

M <- cor(numeric_data)

corrplot(M, 
         method = "color", 
         addCoef.col = "black",
         number.cex = 0.7, 
         tl.col = "black", 
         tl.srt = 45, 
         diag = FALSE)
```


The correlation matrix confirms the improved effectiveness of the GDP variable by taking it into logarithms – an increase from 0.74 to 0.87. An interesting relationship is the number of children a woman bears and life expectancy – having many children indicates a shorter lifespan, but this is most likely conditioned by the socio-economic situation – having many children is most common in developing countries, where people statistically have shorter lifespans. However, the matrix revealed a strong correlation between the infrastructure variables (both among themselves and in relation to GDP). This may indicate collinearity. This compels us to conduct detailed analysis in this area in the future, particularly the use of VIF in the modeling stage.


At the end for this section we are going to analyze how life expectancy varies across continents.
```{r}
boxplot(LifeExp ~ Continent, data = df_final, 
        col = c("gold", "tomato", "skyblue", "lightgreen", "violet"),
        main = "Life Expectancy by Continent",
        ylab = "Years", xlab = "Continent")


ggplot(df_final, aes(x = log_GDP, y = LifeExp, color = Continent)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_minimal() +
  labs(title = "Life Expectancy vs Log(GDP) by Continent")
```


Clear differences between continents are visible. Europe clearly dominates with the highest and highly concentrated results. Africa fares by far the worst. Africa and Asia have very large whiskers, indicating their strong internal differentiation. Analysis of the graphs indicates that the "Continent" variable must be included in the model as a quantitative variable. Without it, the model would distort the results because it would not know that belonging to Europe statistically increases longevity.

#Model preparation- Train/Test split


After the initial analysis of variables, we divide the data into two sets - we will train our model on one set, and then test its effectiveness on the "new" one. We divide the data in an 80/20 ratio.

```{r}
set.seed(123)

train_indices <- sample(1:nrow(df_final), size = 0.8 * nrow(df_final))

train_data <- df_final[train_indices, ]
test_data  <- df_final[-train_indices, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size: ", nrow(test_data))
```

Now let's choose the optimal set of variables, for this task we will use a Hellwig method, which evaluates the explanatory power of each variable in relation to the dependent variable and the redundancy among the explanatory variables. The method calculates a "quality index" for each variable, which helps us identify which variables contribute most to explaining the variance in life expectancy while minimizing redundancy.

```{r}

# Calculate the correlation matrix for the numeric variables in the training dataset, the factorial varaibles will be includes in the model as countries definitely have impact on the life expectancy.

numeric_data <- train_data %>% select(where(is.numeric), -GDP) # usuwamy surowe GDP, zostawiamy log_GDP

cor_matrix <- cor(numeric_data)

r0 <- cor_matrix["LifeExp", -which(colnames(cor_matrix) == "LifeExp")]
r <-cor_matrix[-which(rownames(cor_matrix) == "LifeExp"), -which(colnames(cor_matrix) == "LifeExp")]

scenarios <- expand.grid(rep(list(c(TRUE, FALSE)), ncol(r) )) 
H_final<-c()

for(i in 1:nrow(scenarios) ) {
  indices<-which(as.logical(scenarios[i,]))
  if(length(indices)==0) {
    H_final[i]<-0
    next
  }
  H<-c()
  for(j in 1:length(indices)) {
     H[j]<- r0[indices[j]]^2 / sum(abs(r[indices[j],indices]))
  }
  H_final[i]<-sum(H)
}


best_idx <- which.max(H_final)

best_vars <- colnames(r)[as.logical(scenarios[best_idx, ])]

print(best_vars)

```
# Now let's see the best dataset for our model.

```{r}
print(best_vars)
```


## 4. Model Estimation and Comparison

We decided to compare two approaches: a parsimonious model suggested by the Hellwig method and a full research model that includes all hypothesized predictors.

```{r model_comparison}
# Model 1: Based on Hellwig's selection + factors as country 
model_hellwig <- lm(LifeExp ~ log_GDP + Fertility + Continent, data = train_data)

# Model 2: Full research model (including lifestyle and continents)
# We exclude raw GDP to avoid perfect collinearity with log_GDP
model_full <- lm(LifeExp ~ log_GDP + Fertility + Obesity + Smoking + 
                 Internet + Electricity + CleanFuels + Continent, 
                 data = train_data)

# Reviewing results
summary(model_hellwig)
summary(model_full)
```


The initial estimation of the full model yields a high Adjusted R-squared (0.8152), indicating that our variables collectively explain about 81.5% of the variance in global life expectancy. As expected, log_GDP and Fertility are highly significant ($p < 0.001$), and the Continent dummy variables confirm significant regional differences (e.g., a strong positive effect for Europe and Oceania compared to the African baseline). However, the infrastructure variables (Internet, Electricity, CleanFuels) and lifestyle factors (Smoking, Obesity) lack statistical significance. In econometric modeling, such a scenario—where the overall model fit is high but individual predictors are insignificant—strongly suggests the presence of multicollinearity. To diagnose this, we must examine the Variance Inflation Factor (VIF).


```{r}
vif(model_full)
```


The VIF diagnostic test clearly explains the lack of significance in our previous model. The VIF values for log_GDP (8.73), CleanFuels (7.40), and Internet (7.23) severely exceed the standard acceptable threshold of 5, indicating extreme multicollinearity. Access to electricity, internet, and clean fuels is so strongly coupled with a country's wealth that including them alongside GDP artificially inflates the standard errors of our estimators.

From a scientific standpoint, this compels us to reject Hypothesis 3 (H3): at the macroeconomic level, basic infrastructure does not act as an independent driver of longevity, but is rather a direct function of economic development (log_GDP). Consequently, to establish a robust and mathematically sound model, we will remove the redundant infrastructure variables (Internet, Electricity, CleanFuels) and estimate a reduced model. This will allow us to properly test our remaining hypotheses (H1 and H2) without the distortion of variance inflation.

```{r}
model_final <- lm(LifeExp ~ log_GDP + Fertility + Obesity + Smoking + 
                Continent, 
                 data = train_data)

# Reviewing results

summary(model_final)
```
## 5. Model Diagnostics
Now let's focus on the model, and see whether there is a need to aplicate some changes or not.
```{r}
bptest(model_final)
resettest(model_final, power = 2:3, type = "fitted")
jarque.bera.test(residuals(model_final))
```

As we can see the p-value in RESET test is significanly belov thereshold 0,05, now we will try to implement different models and see if there is a chance for an improvement

```{r}
model_final1 <- lm(log(LifeExp) ~ log_GDP + Fertility + Obesity + Smoking + 
                Continent, 
                 data = train_data)


model_final2 <- lm(LifeExp ~ log_GDP + log(Fertility) + log(Obesity) + log(Smoking) + 
                Continent, 
                 data = train_data)

model_final3 <- lm(LifeExp ~ I(log_GDP^2) + Fertility + Obesity + Smoking + 
                 Continent, 
                 data = train_data)

resettest(model_final1, power = 2:3, type = "fitted")
bptest(model_final1)
jarque.bera.test(residuals(model_final1))
resettest(model_final2, power = 2:3, type = "fitted")
bptest(model_final2)
jarque.bera.test(residuals(model_final2))
resettest(model_final3, power = 2:3, type = "fitted")
bptest(model_final3)
jarque.bera.test(residuals(model_final3))
resettest(model_hellwig, power = 2:3, type = "fitted")
bptest(model_hellwig)
jarque.bera.test(residuals(model_hellwig))

```
```{r}
plot(model_final2, which = 4)
train_data[c(20, 106, 125), ]
model_no_outliers <- lm(LifeExp ~ log_GDP + log(Fertility) + log(Obesity) + log(Smoking) + Continent, 
                        data = train_data[-c(20, 106, 125), ])

jarque.bera.test(residuals(model_no_outliers))

par(mfrow = c(2, 2)) # Dzieli okno graficzne na 4 części
plot(model_final2)
```
# Justification for Removing Outliers

Based on the Cook’s distance plot presented above, several observations were identified as highly influential in the estimation of the regression model.

In particular, observations with indices 20, 106, and 125 exhibit significantly higher Cook’s distance values compared to the rest of the sample. This indicates that these data points have a disproportionately large impact on the estimated coefficients.

A closer inspection reveals that these observations correspond to countries with extreme socio-economic characteristics:

- **Nauru** – exceptionally high obesity and smoking rates,
- **Nigeria** – very high fertility rate and relatively low life expectancy,
- **Lesotho** – unusually low life expectancy given its explanatory variables.

Such observations may violate the standard assumptions of the linear regression model and distort the estimation results by exerting excessive influence on parameter values.

From an econometric perspective, it is justified to temporarily remove these outliers in order to:

- reduce the impact of high-leverage observations,
- improve the distributional properties of residuals,
- assess the robustness of the model.

After removing these observations, the Jarque-Bera test indicates no significant deviation from normality (p-value = 0.6786), which confirms a substantial improvement in model adequacy.

It should be emphasized that the removal of outliers serves as a robustness check rather than a permanent exclusion of data from the analysis.



## 6. Prediction and Error Analysis
In this section, we evaluate the predictive performance of our model using the test dataset. 
We calculate standard error metrics such as MAE, RMSE, and MAPE.

```{r}
predicted<-predict(model_final2, newdata=test_data)
test_residuals<-predicted - test_data$LifeExp
MAE<-mean(abs(test_residuals))
RMSE<-sqrt(mean(test_residuals^2))
MAE
RMSE
MAPE <- mean(abs(test_residuals / test_data$LifeExp)) * 100
MAPE
##############################################################
model_hellwig
predicted<-predict(model_hellwig, newdata=test_data)
test_residuals<-predicted - test_data$LifeExp
MAE<-mean(abs(test_residuals))
RMSE<-sqrt(mean(test_residuals^2))
MAE
RMSE
MAPE <- mean(abs(test_residuals / test_data$LifeExp)) * 100
MAPE

```

## 7. Summary and Interpretation

This study aimed to identify the key determinants of life expectancy across countries using cross-sectional data for 2022. The analysis combined economic, lifestyle, and infrastructure variables to test three main hypotheses regarding the drivers of longevity.

### Model Comparison and Final Choice

Based on both diagnostic tests and predictive performance, we compare the Hellwig model and the extended model (model_final2).

From the diagnostic perspective:

- The Breusch–Pagan test indicates no heteroskedasticity in both models (p-values > 0.05).
- However, the RESET test suggests misspecification in the Hellwig model (p-value = 0.004), while model_final2 does not show this issue (p-value = 0.29).
- The Jarque–Bera test indicates non-normality of residuals in both models, although this problem is mitigated after removing outliers.

From the predictive performance perspective:

 model_final2:
- MAE = 2.86  
- RMSE = 3.40  
- MAPE = 4.01%

 Hellwig model:
- MAE = 2.81  
- RMSE = 3.38  
- MAPE = 3.96%

Although the Hellwig model achieves slightly better predictive accuracy, the differences are very small.

### Final Conclusion – Preferred Model

Taking into account both statistical diagnostics and theoretical considerations, model_final2 is selected as the better model.

Despite slightly worse predictive performance, model_final2 does not suffer from functional form misspecification (RESET test), unlike the Hellwig model. This means that it more accurately captures the underlying relationships between variables.

Additionally, model_final2 includes theoretically justified variables (such as lifestyle factors in logarithmic form), which improves its interpretability and consistency with economic theory.

In contrast, the Hellwig model, although simpler and slightly more accurate in prediction, is misspecified and therefore less reliable for inference.

### Overall Interpretation

The results confirm the main hypotheses:

- H1 (economic development): confirmed  
  GDP per capita (in logarithmic form) has a strong positive impact on life expectancy, with diminishing marginal effects.

- H2 (lifestyle factors): partially confirmed  
  Obesity and smoking have a negative effect, although their statistical significance is limited.

- H3 (infrastructure): rejected 
  Infrastructure variables are not significant once GDP is included, suggesting that their impact is largely captured by economic development.













